Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-04-15 main.tex
Added citation to the preprint and fixed Wegman reference.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles are not self-stable and can be
challenging for the rider to balance. Low speed crashes may be reduced if the
bicycle was self-stable at these speeds by relieving the rider of some of their
necessary control activity. Bicycles can be mechanically modified to lower the
speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
has been known that applying a motor actuated steering torque proportional to
the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
down to very low speeds~\citep{Ruijs1986}. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like natural high speed
self-stability already does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \citet{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

In this paper, we investigate whether an automatically controlled bicycle, that
is stable in a large low speed range for balance assist, is beneficial in
helping to prevent the rider from falling. We test this possible benefit by
applying varying magnitude mechanical perturbations to the handlebars while the
rider is cycling on a treadmill. We assess the difference in the rider's
probability of falling with the balance assist system on and off.

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024a}. At low speeds,
from start up to typical cruising speeds, bicycles are not self-stable and can
be challenging for the rider to balance. Low speed crashes may be reduced if
the bicycle was self-stable at these speeds by relieving the rider of some of
their necessary control activity. Bicycles can be mechanically modified to
lower the speeds at which they are self-stable~\citep{Astrom2005} and since the
1980s it has been known that applying a motor actuated steering torque
proportional to the vehicle's roll angular rate~\footnote{Roll angular rate is
defined herein as per the SAE J670 vehicle standards.} can stabilize a
single-track vehicle down to very low speeds~\citep{Ruijs1986}. If automatic
control of steering can stabilize a bicycle, it may reduce the control required
from the rider to successfully manage balancing tasks much like natural high
speed self-stability already does. We have developed a balance-assisting
bicycle, Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \citet{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

In this paper~\footnote{Also available as a preprint \cite{Haitjema2025}.}, we
investigate whether an automatically controlled bicycle, that is stable in a
large low speed range for balance assist, is beneficial in helping to prevent
the rider from falling. We test this possible benefit by applying varying
magnitude mechanical perturbations to the handlebars while the rider is cycling
on a treadmill. We assess the difference in the rider's probability of falling
with the balance assist system on and off.
